### 2021-02-23 ch/2-related-work.tex
Rearrange things a lot

This work was extended in \citet{calders2009building},
where the same discrimination criterion is used as before,
but called \emph{dependency} between \(s\) and \(y\).
They again present the same dataset changing technique
and a different technique based on giving training examples different weights
instead of re-labeling them.
Examples with \(y=1\) and \(s=0\) get higher weight than those with \(y=0\) and \(s=0\).
For \(s=1\), those with \(y=0\) get higher weight than those with \(y=1\).
Weighting means that when the training examples are sampled from the dataset,
those with higher weight are used more often.
The two approaches are compared on the Adult/Census Income dataset \citep{kohavi1996scaling}
where the task is to estimate if someone earns more than \$50K per year.
The sensitive attribute is \emph{sex}.
In addition to Na\"ive Bayes,
two classifiers based on nearest neighbors and a decision tree classifier are used.
The comparison shows that changing the labels leads to lower dependency (discrimination)
than reweighting the examples.
However, reweighting the examples leads to lower dependency than the baseline.
The difference in accuracy is minor.

@@

\begin{enumerate}
\item
  Prejudice:
  a statistical dependency between the sensitive attribute
  and the class label or the (non-sensitive) input features.
  If the dependency is with the class label, then the paper calls it \emph{indirect prejudice}.
  If it is with the non-sensitive features, they call it \emph{latent prejudice}.
  In line with \citet{pedreshi2008discrimination},
  \emph{direct prejudice} refers to classifiers that make direct use of the sensitive attribute.
  Indirect prejudice and latent prejudice can be quantified by
  the mutual information between \(y\) and \(s\) or \(x\) and \(s\) respectively.
  \citet{kamishima2011fairness} do not consider latent prejudice to be immediately harmful,
  but it can be a problem with respect to the protection of personal data and compliance with laws.
\item
  Underestimation:
  non-convergence of machine learning algorithm due to limited amount of data.
  This magnitude of this problem can be estimated
  by considering the difference between the actual training sample distribution
  and the distribution that the machine learning model has internalized.
  This is the underlying cause of models that make predictions
  that are even more unfair than the dataset.
\item
  Negative Legacy:
  sampling bias and wrong labels in the dataset.
  In contrast to the problem of \emph{prejudice},
  \emph{negative legacy} might not be detectable by analyzing the dataset.
  \emph{Sampling bias}, meaning that certain data points simply are missing,
  and \emph{wrong labels} can only be corrected if other sources of information are available.
\end{enumerate}

We see that previous works nearly exclusively dealt with the first of these issues, \emph{prejudice},
and more precisely \emph{indirect prejudice}.
\citet{kamishima2011fairness} also focus on \emph{indirect prejudice};
essentially because it is the easiest to deal with, apart from \emph{direct prejudice}.
They use \href{https://en.wikipedia.org/wiki/Logistic_regression}{logistic regression} as the basis for their fair classifier.
Their approach to enforcing fairness is qualitatively different from the previous proposals.
Instead of manipulating the dataset or using explicit algorithm to make a classifier fair,
\citet{kamishima2011fairness} treat the fairness constraint like a regularizer.
A term is added to the objective function
that estimates the mutual information between \(y\) and \(s\)
and gets minimized together with the other parts of the objective function.
This strategy leaves most of the work to the optimization that would need to be run anyway.
A factor in front of the regularization term determines how much to value fairness over accuracy.

@@

\subsection{Fairness based on similarity}%
\label{fairness-based-on-similarity}
Nearly all previously discussed papers express some dissatisfaction
with the \emph{Demographic Parity} metric for measuring discrimination.
\citet{luong2011k} use a very different fairness metric
based on the idea that similar people should be treated similarly.
They first define a distance metric which defines a neighborhood
and is basically the Manhattan distance on the z-scores of the attributes.
The distance metric is supposed to be only make use of legally admissible attributes.
An individual is then considered to be unfairly treated
if it is treated differently than its neighbors.
More concretely, for any data point we can check
how many of the \(k\) nearest neighbors have the same class label as that data point.
If the percentage is under a certain threshold
then there was discrimination against the individual corresponding to that data point.
In order to create a fair classifier, for this definition of fairness,
the authors propose to pre-process the dataset,
similar to \citet{kamiran2009classifying} before,
by flipping the class labels of those data points where the class label is considered unfair.
The experiments in the paper show that this method is successful in reducing the discrimination
(according to the given criterion)
for a range of classifiers on the Adult/Census Income dataset.

@@

For these reasons, \citet{dwork2012fairness} try to do better with a metric
that is centered around a measure of similarity between \emph{individuals}.
The goal with this is to achieve \emph{individual fairness} instead of just \emph{group fairness}.
Just like \citet{luong2011k}, they want to treat similar people similarly.
However, the authors do not give a general recipe to construct the necessary similarity measure.
This is left to be decided on a case-by-case basis.
In order for the results to be meaningful,
the similarity measure has to be as close to the ground truth as possible.

The proposed idea is then as follows: a classifier is fair
if and only if the predictive distributions for any two data points
are at least as similar as the two points themselves,
according to a given similarity measure for distributions
and a given similarity measure for data points.
They call this condition the \emph{Lipschitz condition}.
The authors propose two practical similarity measures for distributions
that are well-known in the respective literature.
In order to train a fair classifier,
the Lipschitz condition is then used as a constraint for the optimization.
This is similar to how \citet{kamishima2011fairness} used the fairness criterion as a regularizer.
However, \citet{dwork2012fairness} intend to realize it as a strict constraint.
Furthermore, they show that their criterion is compatible with Demographic Parity
but does not guarantee it in general.
When the two demographic groups are similar according to the individual similarity measure,
then the Lipschitz condition will lead to a good approximation of Demographic Parity.
If the demographic groups are very different,
then the Lipschitz criterion is essentially incompatible with Demographic Parity.

@@

\section{Fair representation}\label{fair-representation}
\citet{zemel2013learning} expand on the idea of creating a \emph{fair representation}.
That is, the features of the data points are transformed such that all bias is removed
but still carry the information necessary for predicting the class label.
This is different from \citet{kamiran2009classifying} where the dataset was manipulated as well,
but the change was restricted to the \emph{labels}.
In this work, the \emph{inputs} are transformed.
This is the first of many works with such a goal.
More specifically, the stated goal of \citet{zemel2013learning} is
to achieve both \emph{group fairness} (in the sense of Demographic Parity)
and \emph{individual fairness} (in the sense of treating similar people similarly),
as \citet{dwork2012fairness} did before.
The learned fair representation is denoted by \(z\).
The condition for fairness is then
\begin{align}
  \label{eq:fair-representation}
  P(s=0|z=z') = P(s=1|z=z') ~.
\end{align}
\citet{zemel2013learning} propose to map the biased inputs \(x\) to \emph{prototypes} in the same space as \(x\).
The probability for being assigned to a particular prototype
must be the same for inputs with \(s=0\) and for \(s=1\).
Based on a given distance function (or similarity measure),
inputs are more likely to be assigned to prototypes that are close-by.
The classification task is then based entirely on the prototypes (the fair representation \(z\)).
In the paper, a linear model is used to map the prototypes to the outcomes \(y\).
Aside from the distance function, the whole model has therefore only two kinds of parameters:
the locations of the prototypes and the weights in the linear model.
These parameters are all optimized together.
The objective function consists of three terms:
the first enforcing Demographic Parity via the prototype locations,
the second ensuring that the prototypes are close to the inputs
and the third targeting a good accuracy by adapting the weights of the classifier.

@@

\citet{feldman2015certifying} try to improve upon the work by \citet{zemel2013learning},
focusing on fair representation as well.
The authors try to ground their definition of fairness in U.S. law.
They define \emph{Disparate Impact} (DI) as
\begin{align}
  \label{eq:disparate-impace}
  DI = \frac{P(y=1|s=0)}{P(y=1|s=1)} ~.
\end{align}
With a target of \(DI = 1\), this would simply enforce Demographic Parity.
However, based on some rulings and recommendations of the U.S. legal system,
they advocate the 80\% Rule, which states that \(DI\) should not be below 80\% (or above 125\%).
This means that the acceptance rate of a disadvantaged demographic group
should not be less than 80\% of the acceptance rate of the other group.
With this definition, there is an explicit allowed range for small unfairness.
Previously, people just tried to get as close as possible to \(DI = 1\),
without stating how close is close enough.

@@

While the algorithms by \citet{feldman2015certifying} are manually constructed and explicit,
\citet{louizos2016variational} use an approach
that falls more in the area of end-to-end learning where fewer hand-crafted algorithms are used.
The method is based on deep variational autoencoders (VAE).
There is an encoder and a decoder that are both modeled as deep neural networks.
The encoder produces the distribution of the latent (fair) representation \(z\)
from the original features \(x\) and the sensitive attribute \(s\).
The decoder recovers the distribution of \(x\) from \(z\) and \(s\).
By choosing a factorized prior \(P(s)P(z)\), a separation between \(s\) and \(z\) is encouraged.

@@

The next question is whether we can achieve all of these criteria simultaneously.
The authors show that a perfect predictor,
which gives a score of \(P(\hat{y}=1|x) = 1\) to data points with \(y=1\)
and \(P(\hat{y}=0|x) = 1\) to those with \(y=0\),
automatically fulfills all three criteria.
As we saw before, this is not true of Demographic Parity.
Furthermore, the authors show that if the dataset is fair in the sense that it satisfies \(P(y=1|s=0) = P(y=1|s=1)\),
\ie, the base acceptance rate is the same for both classes,
then a ``random'' classifier which assigns that base rate as the prediction score to all data points indiscriminately
also fulfills all three criteria.
For this case, even Demographic Parity is fulfilled.
The paper provides a proof
that those two cases are the only ones that achieve the three presented guarantees simultaneously.
Demographic Parity can only be achieved simultaneously with these in the second scenario.
The general case sits between those extremes:
the predictor is not perfect and the dataset is not fair,
and therefore, these definitions of fairness are not compatible.
Note, however, that criterion 2 and 3 \emph{are} compatible with one another.

@@

\citet{hardt2016equality} define the criteria via statistical independence:
\emph{Equalized Odds} refers to the case where \(\hat{y}\) and \(s\) are independent conditional on \(y\).
This is equivalent to the true positive rate and the false positive rate being the same for all groups.
\emph{Equality of Opportunity} is, as mentioned above,
the case where just the true positive rate is the same for all groups.
This means that \(\hat{y}\) and \(s\) are independent conditional on \(y=1\).
One main advantage that these definitions have over Demographic Parity is that a perfect predictor fulfills them.
Demographic Parity is at odds with achieving perfect accuracy.
Furthermore, for Demographic Parity a classifier might be forced to give a positive prediction to a lot of individuals
where it is really not appropriate.

@@

In a follow up work, \citet{zafar2017fairnesstreatment} adapt their work to enforce Equalized Odds,
as defined in \citet{hardt2016equality}.
(However, it is here referred to as ``Disparate Mistreatment'' instead of Equalized Odds.)
As in the previous work, they define a tractable proxy for this measure.
To this end, they first define a function \(g\) that measures
how well the class label \(y \in \{0, 1\}\) and the predicted score for a data point \(0 \leq f(x) \leq 1\) agree:
\begin{align}
  \label{eq:zafar-constraint-2}
  g(y, x) = \min \left(0, \left(y - \tfrac{1}{2}\right) \cdot \left(f(x) - \tfrac{1}{2}\right)\right)
\end{align}
Subtracting \(\tfrac{1}{2}\) is necessary to center the values around \(0\).
The proxy for Equalized Odds is then the covariance between the sensitive attribute and \(g\).
\(g\) can also be modified to only take into account cases with \(y=1\) which corresponds to Equality of Opportunity.
The paper contains experiments on the ProPublica/COMPAS dataset.
The task for this dataset is to predict
whether a criminal offender committed another misconduct or felony within two years,
based on personal information that includes age, gender, race and past criminal history.
The authors compare the proposed method to an implementation of the algorithm from \citet{hardt2016equality}.
In the Equality of Opportunity task, the approach by \citet{hardt2016equality} performs better
(lower discrimination, similar accuracy).
The better performance is likely at least partially explained
by \citet{hardt2016equality} using the sensitive attribute for predictions.
% The evaluation metric is
% the difference in FPR (false positive rate) and the difference in FNR (false negative rate) between the two groups.
% Hardt et al.~achieve an FPR difference of 0.02 at 66\% accuracy while Zafar et al.~achieve an FPR difference of
% 0.06 at the same 66\% accuracy.
For Equalized Odds,
the proposed approach achieves higher accuracy but also higher discrimination than in \citet{hardt2016equality}.
One limiting factor that the authors identify is the size of the dataset:
the degree of discrimination cannot be accurately estimated by the algorithm if the dataset is small.

\citet{woodworth2017learning} also consider Equalized Odds as a fairness definition
but give strong criticism for the post-processing approach from \citet{hardt2016equality}.
They present examples where finding the optimal (but discriminatory) predictor in a particular hypothesis class
and then correcting it post hoc performs poorly.
The proofs are based on the PAC learning framework.
Certain training distributions allow the training of good fair classifiers from scratch
but when doing post-processing on a Bayes-optimal predictor for this distribution,
the result can be a very bad predictor that gives essentially random predictions.
The authors explicitly construct one such distribution:
it corresponds to the case where the sensitive attribute \(s\) is more predictive of the class label \(y\)
than the features \(x\), which is arguably an unusual case.

@@

\section{Fairness via causal reasoning}%
\label{sec:fairness-definition-based-on-causal-reasoning}
In this section, I will briefly discuss a view that also falls under the definition-centric view:
in it, the goal is to make decisions that conform to causality-based notions of fairness.
Methods in this category assume that the causal structure of the task is known,
which can then be used to identify unfair path ways for making decisions.
One motivating example is this \citep{dedeo2014wrong}:
In college admissions, we might not want to discriminate against prospective students from poorer backgrounds.
If the goal is to admit those students that have a chance of graduating,
then we could employ a machine learning algorithm to predict graduation rate.
One attribute that might turn out to be predictive of graduation is physical fitness.
If this attribute's causal influence is through,
for example, signaling a character trait such as grit,
then this is an admissible attribute.
However, if on the other hand physical fitness is caused by access to expensive gyms,
then it is a signal for high socioeconomic status which we do not want to use as a criterion.
So, depending on what the causal mechanism is,
physical fitness could be a discriminatory or a non-discriminatory feature.

@@

% The difficult problem that Pearl solved with his theory of counterfactuals
% is how to define the counterfactual world where the value of \(s\) is \(j\) instead of \(i\).
% One approach that philosophers took in trying to define it
% is looking for the ``closest world'' to ours where \(s=j\), according to some measure of distance.
% Pearl's counterfactuals do not work this way, but it is a useful intuition.
Making sure predictions are the same, for the closest world with \(s=j\),
is arguably what \citet{dwork2012fairness} (see above) did
with their idea of treating similar people similarly (regardless of \(s\)).
In this way, counterfactual fairness is similar to the individual fairness
that \citet{dwork2012fairness} defined.
However, instead of using a similarity metric,
a causal model is needed for counterfactual fairness.
The two criteria share some strengths and weaknesses.
A strength is that they are both fair on the individual level,
which is not the case for Demographic Parity and Equality of Opportunity.
In Demographic Parity and Equality of Opportunity,
we can achieve fairness by doing ``nege'' discrimination on some individuals
and ``positive'' discrimination on others, as long as those two effects cancel out.
This is because those criteria are defined on \emph{groups} and not individuals.
However, in turn this implies that individual fairness and counterfactual fairness
could not be used to implement affirmative action (``positive'' discrimination for a group)
and they rely on the features \(x\) being correct.

@@

\citet{kim2019learning} also consider the problem of learning from biased data (and evaluating on unbiased data).
While they do not formulate the problem as a fairness problem,
it is equivalent to enforcing fairness in the presence of severe sampling bias.
Their main motivating example is the Colored MNIST dataset:
in this dataset, digits are randomly colored in the test set,
but in the training set there is a one-to-one correspondence between digit class and color.
A na\"ive classifier will learn to predict color instead of digit class.
The proposed method is very similar to \citet{ganin2016domain} and \citet{edwards2016censoring},
if we think of the colors as forming different domains
(\ie, the \emph{red} domain contains all red digits, etc).
The goal is then to learn a domain-independent representation.
The main difference to previous work is an additional entropy loss term in the adversarial loss.

\citet{arjovsky2019invariant} tackle a similar problem, but take a very different approach,
which they term ``Invariant Risk Minimization'', in contrast to the default method Empirical Risk Minimization.
Just like the approach by \citet{kim2019learning}, 
As with \citet{kim2019learning},
the goal is to ignore spurious correlations that only appear in the very imperfect training set,
but not in the test set;
and they also perform experiments on Colored MNIST (see above).
They formalize the problem as one of different environments $e\in E$,
where each environment is a different, biased view of the same underlying data distribution.
The idea is to train a predictor that is simultaneously optimal for all the environments,
with the expectation that this generalizes to the test set.
This idea is formulated as a (untractable) bi-level, constrained optimization problem,
which they approximate with a tractable regularized optimization.

\citet{creager2020environment} build directly on \citet{arjovsky2019invariant}.
They aim to address the limitation that the environments -- that one wants to be invariant to -- have to be pre-defined.
Their contribution is to infer these environments instead,
based on identifying which environment splits would most negatively affect an ERM classifier.
Their experiments show that this can even approve upon hand-crafted environments;
experiments are performed on a synthetic dataset and Colored MNIST.
In contrast to \citet{kim2019learning} and \citet{arjovsky2019invariant},
the authors explicitly point out the connection to traditional fairness methods.

@@

\citet{jiang2020identifying} also formulate their problem this way:
in their setting, the underlying unbiased labels become corrupted by a biased labeler,
which results in a biased training set.
Their goal is to train a classifier that makes correct predictions consistent with the true labels.
However, they treat the true labels as completely unknown
and evaluate their models w.r.t.\ the common fairness definitions on a test set
that is just as biased as the training set.

>>>>

This work was extended in \citet{calders2009building},
where the same discrimination criterion is used as before,
but called \emph{dependency} between \(s\) and \(y\).
They again present the same dataset changing technique
and a different technique based on giving training examples different weights
instead of re-labelling them.
Examples with \(y=1\) and \(s=0\) get higher weight than those with \(y=0\) and \(s=0\).
For \(s=1\), those with \(y=0\) get higher weight than those with \(y=1\).
Weighting means that when the training examples are sampled from the dataset,
those with higher weight are used more often.
The two approaches are compared on the Adult/Census Income dataset \citep{kohavi1996scaling}
where the task is to estimate if someone earns more than \$50K per year.
The sensitive attribute is \emph{sex}.
In addition to Na\"ive Bayes,
two classifiers based on nearest neighbours and a decision tree classifier are used.
The comparison shows that changing the labels leads to lower dependency (discrimination)
than reweighting the examples.
However, reweighting the examples leads to lower dependency than the baseline.
The difference in accuracy is minor.

@@

\begin{enumerate}
\item
  Prejudice:
  a statistical dependency between the sensitive attribute
  and the class label or the (non-sensitive) input features.
  If the dependency is with the class label, then the paper calls it \emph{indirect prejudice}.
  If it is with the non-sensitive features, they call it \emph{latent prejudice}.
  In line with \citet{pedreshi2008discrimination},
  \emph{direct prejudice} refers to classifiers that make direct use of the sensitive attribute.
  Indirect prejudice and latent prejudice can be quantified by
  the mutual information between \(y\) and \(s\) or \(x\) and \(s\) respectively.
  \citet{kamishima2011fairness} do not consider latent prejudice to be immediately harmful,
  but it can be a problem with respect to the protection of personal data and compliance with laws.
\item
  Underestimation:
  non-convergence of machine learning algorithm due to limited amount of data.
  This magnitude of this problem can be estimated
  by considering the difference between the actual training sample distribution
  and the distribution that the machine learning model has internalized.
  This is the underlying cause of models that make predictions
  that are even more unfair than the dataset.
\item
  Negative Legacy:
  sampling bias and wrong labels in the dataset.
  In contrast to the problem of \emph{prejudice},
  \emph{negative legacy} might not be detectable by analysing the dataset.
  \emph{Sampling bias}, meaning that certain data points simply are missing,
  and \emph{wrong labels} can only be corrected if other sources of information are available.
\end{enumerate}

We see that previous works nearly exclusively dealt with the first of these issues, \emph{prejudice},
and more precisely \emph{indirect prejudice}.
\citet{kamishima2011fairness} also focus on \emph{indirect prejudice};
essentially because it is the easiest to deal with, apart from \emph{direct prejudice}.
They use \acf{LR} as the basis for their fair classifier.
Their approach to enforcing fairness is qualitatively different from the previous proposals.
Instead of manipulating the dataset or using explicit algorithm to make a classifier fair,
\citet{kamishima2011fairness} treat the fairness constraint like a regulariser.
A term is added to the objective function
that estimates the mutual information between \(y\) and \(s\)
and gets minimized together with the other parts of the objective function.
This strategy leaves most of the work to the optimization that would need to be run anyway.
A factor in front of the regularization term determines how much to value fairness over accuracy.

@@

\subsection{Fairness based on similarity}%
\label{fairness-based-on-similarity}
Nearly all previously discussed papers express some dissatisfaction
with the \emph{Demographic Parity} metric for measuring discrimination.
\citet{luong2011k} use a very different fairness metric
based on the idea that similar people should be treated similarly.
They first define a distance metric which defines a neighbourhood
and is basically the Manhattan distance on the z-scores of the attributes.
The distance metric is supposed to be only make use of legally admissible attributes.
An individual is then considered to be unfairly treated
if it is treated differently than its neighbours.
More concretely, for any data point we can check
how many of the \(k\) nearest neighbours have the same class label as that data point.
If the percentage is under a certain threshold
then there was discrimination against the individual corresponding to that data point.
In order to create a fair classifier, for this definition of fairness,
the authors propose to pre-process the dataset,
similar to \citet{kamiran2009classifying} before,
by flipping the class labels of those data points where the class label is considered unfair.
The experiments in the paper show that this method is successful in reducing the discrimination
(according to the given criterion)
for a range of classifiers on the Adult/Census Income dataset.

@@

For these reasons, \citet{dwork2012fairness} try to do better with a metric
that is centred around a measure of similarity between \emph{individuals}.
The goal with this is to achieve \emph{individual fairness} instead of just \emph{group fairness}.
Just like \citet{luong2011k}, they want to treat similar people similarly.
However, the authors do not give a general recipe to construct the necessary similarity measure.
This is left to be decided on a case-by-case basis.
In order for the results to be meaningful,
the similarity measure has to be as close to the ground truth as possible.

The proposed idea is then as follows: a classifier is fair
if and only if the predictive distributions for any two data points
are at least as similar as the two points themselves,
according to a given similarity measure for distributions
and a given similarity measure for data points.
They call this condition the \emph{Lipschitz condition}.
The authors propose two practical similarity measures for distributions
that are well-known in the respective literature.
In order to train a fair classifier,
the Lipschitz condition is then used as a constraint for the optimization.
This is similar to how \citet{kamishima2011fairness} used the fairness criterion as a regulariser.
However, \citet{dwork2012fairness} intend to realize it as a strict constraint.
Furthermore, they show that their criterion is compatible with Demographic Parity
but does not guarantee it in general.
When the two demographic groups are similar according to the individual similarity measure,
then the Lipschitz condition will lead to a good approximation of Demographic Parity.
If the demographic groups are very different,
then the Lipschitz criterion is essentially incompatible with Demographic Parity.

@@

\section{Fair representation}\label{sec:fair-representation}
\citet{zemel2013learning} expand on the idea of creating a \emph{fair representation}.
That is, the features of the data points are transformed such that all bias is removed
but still carry the information necessary for predicting the class label.
This is different from \citet{kamiran2009classifying} where the dataset was manipulated as well,
but the change was restricted to the \emph{labels}.
In this work, the \emph{inputs} are transformed.
This is the first of many works with such a goal.
More specifically, the stated goal of \citet{zemel2013learning} is
to achieve both \emph{group fairness} (in the sense of Demographic Parity)
and \emph{individual fairness} (in the sense of treating similar people similarly),
as \citet{dwork2012fairness} did before.
The learned fair representation is denoted by \(z\).
The condition for fairness is then
\begin{align}
  \label{eq:fair-representation}
  P(s=0|z=z') = P(s=1|z=z') ~.
\end{align}
\citet{zemel2013learning} propose to map the biased inputs \(x\) to \emph{prototypes} in the same space as \(x\).
The probability for being assigned to a particular prototype
must be the same for inputs with \(s=0\) and for \(s=1\).
Based on a given distance function (or similarity measure),
inputs are more likely to be assigned to prototypes that are close-by.
The classification task is then based entirely on the prototypes (the fair representation \(z\)).
In the paper, a linear model is used to map the prototypes to the outcomes \(y\).
Aside from the distance function, the whole model has therefore only two kinds of parameters:
the locations of the prototypes and the weights in the linear model.
These parameters are all optimized together.
The objective function consists of three terms:
the first enforcing Demographic Parity via the prototype locations,
the second ensuring that the prototypes are close to the inputs
and the third targeting a good accuracy by adapting the weights of the classifier.

@@

\citet{feldman2015certifying} try to improve upon the work by \citet{zemel2013learning},
focusing on fair representation as well.
The authors try to ground their definition of fairness in U.S.\ law.
They define \emph{Disparate Impact} (DI) as
\begin{align}
  \label{eq:disparate-impace}
  DI = \frac{P(y=1|s=0)}{P(y=1|s=1)} ~.
\end{align}
With a target of \(DI = 1\), this would simply enforce Demographic Parity.
However, based on some rulings and recommendations of the U.S.\ legal system,
they advocate the 80\% Rule, which states that \(DI\) should not be below 80\% (or above 125\%).
This means that the acceptance rate of a disadvantaged demographic group
should not be less than 80\% of the acceptance rate of the other group.
With this definition, there is an explicit allowed range for small unfairness.
Previously, people just tried to get as close as possible to \(DI = 1\),
without stating how close is close enough.

@@

While the algorithms by \citet{feldman2015certifying} are manually constructed and explicit,
\citet{louizos2016variational} use an approach
that falls more in the area of end-to-end learning where fewer hand-crafted algorithms are used.
The method is based on deep variational autoencoders (VAE).
There is an encoder and a decoder that are both modelled as deep neural networks.
The encoder produces the distribution of the latent (fair) representation \(z\)
from the original features \(x\) and the sensitive attribute \(s\).
The decoder recovers the distribution of \(x\) from \(z\) and \(s\).
By choosing a factorized prior \(P(s)P(z)\), a separation between \(s\) and \(z\) is encouraged.

@@

The next question is whether we can achieve all of these criteria simultaneously.
The authors show that a perfect predictor,
which gives a score of \(P(\hat{y}=1|x) = 1\) to data points with \(y=1\)
and \(P(\hat{y}=0|x) = 1\) to those with \(y=0\),
automatically satisfies all three criteria.
As we saw before, this is not true of Demographic Parity.
Furthermore, the authors show that if the dataset is fair in the sense that it satisfies \(P(y=1|s=0) = P(y=1|s=1)\),
\ie, the base acceptance rate is the same for both classes,
then a ``random'' classifier which assigns that base rate as the prediction score to all data points indiscriminately
also satisfies all three criteria.
For this case, even Demographic Parity is fulfilled.
The paper provides a proof
that those two cases are the only ones that achieve the three presented guarantees simultaneously.
Demographic Parity can only be achieved simultaneously with these in the second scenario.
The general case sits between those extremes:
the predictor is not perfect and the dataset is not fair,
and therefore, these definitions of fairness are not compatible.
Note, however, that criterion 2 and 3 \emph{are} compatible with one another.

@@

\citet{hardt2016equality} define the criteria via statistical independence:
\emph{Equalized Odds} refers to the case where \(\hat{y}\) and \(s\) are independent conditional on \(y\).
This is equivalent to the true positive rate and the false positive rate being the same for all groups.
\emph{Equality of Opportunity} is, as mentioned above,
the case where just the true positive rate is the same for all groups.
This means that \(\hat{y}\) and \(s\) are independent conditional on \(y=1\).
One main advantage that these definitions have over Demographic Parity is that a perfect predictor satisfies them.
Demographic Parity is at odds with achieving perfect accuracy.
Furthermore, for Demographic Parity a classifier might be forced to give a positive prediction to a lot of individuals
where it is really not appropriate.

@@

In a follow up work, \citet{zafar2017fairnesstreatment} adapt their work to enforce Equalized Odds,
as defined in \citet{hardt2016equality}.
(However, it is here referred to as ``Disparate Mistreatment'' instead of Equalized Odds.)
As in the previous work, they define a tractable proxy for this measure.
To this end, they first define a function \(g\) that measures
how well the class label \(y \in \{0, 1\}\) and the predicted score for a data point \(0 \leq f(x) \leq 1\) agree:
\begin{align}
  \label{eq:zafar-constraint-2}
  g(y, x) = \min \left(0, \left(y - \tfrac{1}{2}\right) \cdot \left(f(x) - \tfrac{1}{2}\right)\right)
\end{align}
Subtracting \(\tfrac{1}{2}\) is necessary to centre the values around \(0\).
The proxy for Equalized Odds is then the covariance between the sensitive attribute and \(g\).
\(g\) can also be modified to only take into account cases with \(y=1\) which corresponds to Equality of Opportunity.
The paper contains experiments on the ProPublica/COMPAS dataset.
The task for this dataset is to predict
whether a criminal offender committed another misconduct or felony within two years,
based on personal information that includes age, gender, race and past criminal history.
The authors compare the proposed method to an implementation of the algorithm from \citet{hardt2016equality}.
In the Equality of Opportunity task, the approach by \citet{hardt2016equality} performs better
(lower discrimination, similar accuracy).
The better performance is likely at least partially explained
by \citet{hardt2016equality} using the sensitive attribute for predictions.
% The evaluation metric is
% the difference in FPR (false positive rate) and the difference in FNR (false negative rate) between the two groups.
% Hardt et al.~achieve an FPR difference of 0.02 at 66\% accuracy while Zafar et al.~achieve an FPR difference of
% 0.06 at the same 66\% accuracy.
For Equalized Odds,
the proposed approach achieves higher accuracy but also higher discrimination than in \citet{hardt2016equality}.
One limiting factor that the authors identify is the size of the dataset:
the degree of discrimination cannot be accurately estimated by the algorithm if the dataset is small.

\citet{woodworth2017learning} also consider Equalized Odds as a fairness definition
but give strong criticism for the post-processing approach from \citet{hardt2016equality}.
They present examples where finding the optimal (but discriminatory) predictor in a particular hypothesis class
and then correcting it post-hoc performs poorly.
The proofs are based on the PAC learning framework.
Certain training distributions allow the training of good fair classifiers from scratch
but when doing post-processing on a Bayes-optimal predictor for this distribution,
the result can be a very bad predictor that gives essentially random predictions.
The authors explicitly construct one such distribution:
it corresponds to the case where the sensitive attribute \(s\) is more predictive of the class label \(y\)
than the features \(x\), which is arguably an unusual case.

@@

\section{Fairness via causal reasoning}%
\label{sec:fairness-definition-based-on-causal-reasoning}
In this section, I will briefly discuss a view that also falls under the definition-centric view:
in it, the goal is to make decisions that conform to causality-based notions of fairness.
Methods in this category assume that the causal structure of the task is known,
which can then be used to identify unfair path ways for making decisions.
One motivating example is this \citep{dedeo2014wrong}:
In college admissions, we might not want to discriminate against prospective students from poorer backgrounds.
If the goal is to admit those students that have a chance of graduating,
then we could employ a machine learning algorithm to predict graduation rate.
One attribute that might turn out to be predictive of graduation is physical fitness.
If this attribute's causal influence is through,
for example, signalling a character trait such as grit,
then this is an admissible attribute.
However, if on the other hand physical fitness is caused by access to expensive gyms,
then it is a signal for high socioeconomic status which we do not want to use as a criterion.
So, depending on what the causal mechanism is,
physical fitness could be a discriminatory or a non-discriminatory feature.

@@

% The difficult problem that Pearl solved with his theory of counterfactuals
% is how to define the counterfactual world where the value of \(s\) is \(j\) instead of \(i\).
% One approach that philosophers took in trying to define it
% is looking for the ``closest world'' to ours where \(s=j\), according to some measure of distance.
% Pearl's counterfactuals do not work this way, but it is a useful intuition.
Making sure predictions are the same, for the closest world with \(s=j\),
is arguably what \citet{dwork2012fairness} (see above) did
with their idea of treating similar people similarly (regardless of \(s\)).
In this way, counterfactual fairness is similar to the individual fairness
that \citet{dwork2012fairness} defined.
However, instead of using a similarity metric,
a causal model is needed for counterfactual fairness.
The two criteria share some strengths and weaknesses.
A strength is that they are both fair on the individual level,
which is not the case for Demographic Parity and Equality of Opportunity.
In Demographic Parity and Equality of Opportunity,
we can achieve fairness by doing ``negative'' discrimination on some individuals
and ``positive'' discrimination on others, as long as those two effects cancel out.
This is because those criteria are defined on \emph{groups} and not individuals.
However, in turn this implies that individual fairness and counterfactual fairness
could not be used to implement affirmative action (``positive'' discrimination for a group)
and they rely on the features \(x\) being correct.

@@

\citet{kim2019learning} also consider the problem of learning from biased data (and evaluating on unbiased data).
While they do not formulate the problem as a fairness problem,
it is equivalent to enforcing fairness in the presence of severe sampling bias.
Their main motivating example is the Coloured MNIST dataset:
in this dataset, digits are randomly coloured in the test set,
but in the training set there is a one-to-one correspondence between digit class and colour.
A na\"ive classifier will learn to predict colour instead of digit class.
The proposed method is very similar to \citet{ganin2016domain} and \citet{edwards2016censoring},
if we think of the colours as forming different domains
(\ie, the \emph{red} domain contains all red digits, etc).
The goal is then to learn a domain-independent representation.
The main difference to previous work is an additional entropy loss term in the adversarial loss.

\citet{arjovsky2019invariant} tackle a similar problem, but take a very different approach,
which they term ``Invariant Risk Minimization'', in contrast to the default method Empirical Risk Minimization.
Just like the approach by \citet{kim2019learning}, 
As with \citet{kim2019learning},
the goal is to ignore spurious correlations that only appear in the very imperfect training set,
but not in the test set;
and they also perform experiments on Coloured MNIST (see above).
They formalize the problem as one of different environments $e\in E$,
where each environment is a different, biased view of the same underlying data distribution.
The idea is to train a predictor that is simultaneously optimal for all the environments,
with the expectation that this generalizes to the test set.
This idea is formulated as a (intractable) bi-level, constrained optimization problem,
which they approximate with a tractable regularized optimization.

\citet{creager2020environment} build directly on \citet{arjovsky2019invariant}.
They aim to address the limitation that the environments -- that one wants to be invariant to -- have to be pre-defined.
Their contribution is to infer these environments instead,
based on identifying which environment splits would most negatively affect an ERM classifier.
Their experiments show that this can even approve upon hand-crafted environments;
experiments are performed on a synthetic dataset and Coloured MNIST.
In contrast to \citet{kim2019learning} and \citet{arjovsky2019invariant},
the authors explicitly point out the connection to traditional fairness methods.

@@

\citet{jiang2020identifying} also formulate their problem this way:
in their setting, the underlying unbiased labels become corrupted by a biased labeller,
which results in a biased training set.
Their goal is to train a classifier that makes correct predictions consistent with the true labels.
However, they treat the true labels as completely unknown
and evaluate their models w.r.t.\ the common fairness definitions on a test set
that is just as biased as the training set.
